Evaluation of Diagnostic Concordance Between Algorithms for Parkinson’s Disease Dementia

### 2025-10-12 _manuscript/manuscript.qmd

---
lang: en
title: "Evaluation of Diagnostic Concordance Between Algorithms for Parkinson’s Disease Dementia"
shorttitle: "Parkinson's Disease Dementia Level I Criteria"
author:
  - name: Martina Mana
    corresponding: false
    orcid: "0009-0007-4665-3946"
    role:
      - Conceptualization
      - Data curation
      - Writing - original draft
    affiliations:
      - id: "1FM"
        name: "First Faculty of Medicine and General University Hospital in Prague, Charles University, Czech Republic"
        department: "Department of Neurology and Centre of Clinical Neuroscience"
  - name: Josef Mana
    corresponding: false
    orcid: "0000-0002-7817-3978" 
    email: "josef.mana@lf1.cuni.cz"
    role:
      - Conceptualization
      - Data curation
      - Investigation
      - Formal analysis
      - Software
      - Methodology
      - Project administration
      - Validation
      - Writing - original draft
    affiliations:
      - id: "1FM"
        name: "First Faculty of Medicine and General University Hospital in Prague, Charles University, Czech Republic"
        department: "Department of Neurology and Centre of Clinical Neuroscience"
  - name: Tereza Uhrova
    corresponding: false
    email: "tereza.uhrova@vfn.cz"
    role:
      - Investigation
    affiliations:
      - id: "1FM"
        name: "First Faculty of Medicine and General University Hospital in Prague, Charles University, Czech Republic"
        department: "Department of Neurology and Centre of Clinical Neuroscience"
  - name: Robert Jech
    corresponding: false
    orcid: "0000-0002-9732-8947"
    email: "jech@cesnet.cz"
    role:
      - Funding acquisition
      - Resources
      - Writing - review & editing
    affiliations:
      - id: "1FM"
        name: "First Faculty of Medicine and General University Hospital in Prague, Charles University, Czech Republic"
        department: "Department of Neurology and Centre of Clinical Neuroscience"
  - name: Ondrej Bezdicek
    corresponding: true
    orcid: "0000-0002-5108-0181"
    email: "ondrej.bezdicek@gmail.com"
    role:
      - Investigation
      - Data curation
      - Funding acquisition
      - Conceptualization
      - Project administration
      - Supervision
      - Writing - original draft
    affiliations:
      - id: "1FM"
        name: "First Faculty of Medicine and General University Hospital in Prague, Charles University, Czech Republic"
        department: "Department of Neurology and Centre of Clinical Neuroscience"
abstract: ""
keywords:
  - Parkinson’s disease
  - Parkinson's disease dementia
  - level I criteria
format:
 apaquarto-pdf:
    documentmode: man
    keep-tex: true
 apaquarto-docx: 
    toc: false
author-note:
  disclosures:
    financial-support: "Supported by The project National Institute for Neurological Research (Programme EXCELES, ID Project No. LX22NPO5107) - Funded by the European Union – Next Generation EU; Charles University: Cooperatio Program in Neuroscience; General University Hospital in Prague project MH CZ-DRO-VFN64165."
bibliography: bibliography.bib
floatsintext: false
numbered-lines: false
suppress-title-page: false
warning: false
echo: false
---

@@

Using Level II diagnosis as the reference, MoCA-based Level I algorithms, particularly those using Clock Drawing to assess executive function, demonstrated the highest predictive accuracy. This supports recent proposals to modernize PDD diagnostic frameworks [@kulisevsky2024], favouring MoCA-derived components and broader functional assessment tools. In contrast, MMSE-based algorithms consistently underperformed, suggesting limited sensitivity in capturing cognitive deficits typical in PDD.

@@

Our study highlights the variability in PDD classification across Level I diagnostic algorithms, heavily influenced by IADL operationalization and the choice of cognitive screening tools. The findings support the call for a change of the current diagnostic criteria [@kulisevsky2024], favouring the use of MoCA-based components and comprehensive IADL assessments.

>>>>

---
lang: en
title: "Evaluation of Diagnostic Concordance Between Algorithms for Parkinson’s Disease Dementia"
shorttitle: "Parkinson's Disease Dementia Level I Criteria"
author:
  - name: Martina Mana
    corresponding: false
    orcid: "0009-0007-4665-3946"
    role:
      - Conceptualization
      - Data curation
      - Writing - original draft
    affiliations:
      - id: "1FM"
        name: "First Faculty of Medicine and General University Hospital in Prague, Charles University, Czech Republic"
        department: "Department of Neurology and Centre of Clinical Neuroscience"
  - name: Josef Mana
    corresponding: false
    orcid: "0000-0002-7817-3978" 
    email: "josef.mana@lf1.cuni.cz"
    role:
      - Conceptualization
      - Data curation
      - Investigation
      - Formal analysis
      - Software
      - Methodology
      - Project administration
      - Validation
      - Writing - original draft
    affiliations:
      - id: "1FM"
        name: "First Faculty of Medicine and General University Hospital in Prague, Charles University, Czech Republic"
        department: "Department of Neurology and Centre of Clinical Neuroscience"
  - name: Tereza Uhrova
    corresponding: false
    email: "tereza.uhrova@vfn.cz"
    role:
      - Investigation
    affiliations:
      - id: "1FM"
        name: "First Faculty of Medicine and General University Hospital in Prague, Charles University, Czech Republic"
        department: "Department of Neurology and Centre of Clinical Neuroscience"
  - name: Robert Jech
    corresponding: false
    orcid: "0000-0002-9732-8947"
    email: "jech@cesnet.cz"
    role:
      - Funding acquisition
      - Resources
      - Writing - review & editing
    affiliations:
      - id: "1FM"
        name: "First Faculty of Medicine and General University Hospital in Prague, Charles University, Czech Republic"
        department: "Department of Neurology and Centre of Clinical Neuroscience"
  - name: Ondrej Bezdicek
    corresponding: true
    orcid: "0000-0002-5108-0181"
    email: "ondrej.bezdicek@gmail.com"
    role:
      - Investigation
      - Data curation
      - Funding acquisition
      - Conceptualization
      - Project administration
      - Supervision
      - Writing - original draft
    affiliations:
      - id: "1FM"
        name: "First Faculty of Medicine and General University Hospital in Prague, Charles University, Czech Republic"
        department: "Department of Neurology and Centre of Clinical Neuroscience"
abstract: "Background: A significant number of patients with Parkinson’s disease (PD) gradually progress to Parkinson’s disease dementia (PDD). A recent “Call for Change” (Kulisevsky et al., 2024) proposes updates to the current diagnostic criteria for PDD, including the use of different screening tests and broader functional assessments. Objective: We aimed to evaluate the diagnostic concordance between several algorithms for PDD based on Level I (i.e., screening) criteria and to assess their predictive validity for Level II (i.e., neuropsychological battery) diagnosis. Methods: We conducted a cross-sectional retrospective analysis of 204 patients diagnosed with PD. All patients underwent a comprehensive neuropsychological battery. A total of 68 diagnostic algorithms were operationalised based on the combinations of various cognitive assessments, including the Mini Mental State Exam (MMSE), the Montreal Cognitive Assessment (MoCA), and the shortened version of MoCA (sMoCA). Functional impairment was based on the Functional Assessment Questionnaire (FAQ), specifically FAQ item 9 or the FAQ total score. Concordance between algorithms was assessed using pairwise Cohen’s . The predictive validity of Level I algorithms for Level II diagnosis was assessed using Exact Binomial Tests. Results: PDD rate estimates varied widely across algorithms from 2.00% up to 16.75%. Higher rates were associated with functional impairment based on the definition of FAQ score ( = 0.75, = 0.86). Moderate to high concordance was observed among algorithms using the same functional impairment definition. MoCA-based algorithms most accurately predicted Level II classification, especially those using Clock Drawing to assess executive function. Conclusions: Diagnostic outcomes for PDD are sensitive to the choice of cognitive and functional instruments. Our results support some of the proposed changes to PDD diagnostic criteria on Level I, emphasising the consistency of MoCA-based assessments and comprehensive functional impairment evaluation."
keywords:
  - Parkinson’s disease
  - dementia
  - diagnostic criteria
  - functional assessment
  - cognitive assessment
format:
 apaquarto-pdf:
    documentmode: man
    keep-tex: true
 apaquarto-docx: 
    toc: false
author-note:
  disclosures:
    financial-support: "Supported by The project National Institute for Neurological Research (Programme EXCELES, ID Project No. LX22NPO5107) - Funded by the European Union – Next Generation EU; Charles University: Cooperatio Program in Neuroscience; General University Hospital in Prague project MH CZ-DRO-VFN64165."
bibliography: bibliography.bib
floatsintext: false
numbered-lines: false
suppress-title-page: false
warning: false
echo: false
---

@@

Using Level II diagnosis as the gold standard, MoCA-based Level I algorithms,
particularly those using Clock Drawing to assess executive function, demonstrated the highest predictive accuracy. This supports recent proposals to modernise PDD diagnostic frameworks [@kulisevsky2024], favouring MoCA-derived components and more specific, PD-tailored functional assessment tools. In contrast, MMSE-based algorithms consistently underperformed, suggesting limited sensitivity in capturing cognitive deficits typical in PDD.

@@

In sum,our study systematically investigated how varying definitions of impaired cognition and IADL deficit affect PDD diagnostic accuracy. Our study highlights the variability in PDD classification across Level I diagnostic algorithms, significantly influenced by IADL operationalisation and the choice of cognitive screening tools. The findings strongly support the call for a change of the current diagnostic criteria [@kulisevsky2024], favouring the use of MoCA-based components and comprehensive IADL assessments.
